
# Coursera: Financial Engineering II: In-class Exercises in R

* [Lesson 1.1: Mean Variance Overview](#Lesson-1.1:-Mean-Variance-Overview)
* [Lesson 1.3 Sharpe Optimal Portfolio and the Risk-Free Frontier](#Lesson-1.3-Sharpe-Optimal-Portfolio-and-the-Risk-Free-Frontier)

```{r}

    options(width=100)

```

<br /><br />
## <a name="Lesson-1.1:-Mean-Variance-Overview"></a>Lesson 1.1: Mean Variance Overview


```{r}
   

    asset.names <- c("US Bonds",
                     "Int’l Bonds",
                     "US Large Growth",
                     "US Large Value",
                     "US Small Growth",
                     "US Small Value",
                     "Int’l Dev. Equity",
                     "Int’l Emerg. Equity" )

    mu.vec <- c( 3.1500000,
                 1.75000000,
                 -6.39000000,
                 -2.86000000,
                 -6.75000000,
                 -0.54000000,
                 -6.75000000,
                 -5.26000000 ) / 100

    names(mu.vec) <- asset.names
    d <- length(asset.names)            # number of assets

    #
    # covariance matrix 
    #
    Sigma.mat <- matrix(c(0.0010050, 0.00132800, -0.00057900, -0.00067500, 0.00012100, 0.00012800, -0.00044500, -0.00043700,
                          0.0013280, 0.00727700, -0.00130700, -0.00061000, -0.00223700, -0.00098900, 0.00144200, -0.00153500,
                          -0.0005790, -0.00130700, 0.05985200, 0.02758800, 0.06349700, 0.02303600, 0.03296700, 0.04803900,
                          -0.0006750, -0.00061000, 0.02758800, 0.02960900, 0.02657200, 0.02146500, 0.02069700, 0.02985400,
                          0.0001210, -0.00223700, 0.06349700, 0.02657200, 0.10248800, 0.04274400, 0.03994300, 0.06599400,
                          0.0001280, -0.00098900, 0.02303600, 0.02146500, 0.04274400, 0.03205600, 0.01988100, 0.03223500,
                          -0.0004450, 0.00144200, 0.03296700, 0.02069700, 0.03994300, 0.01988100, 0.02835500, 0.03506400,
                          -0.0004370, -0.00153500, 0.04803900, 0.02985400, 0.06599400, 0.03223500, 0.03506400, 0.07995800),
                        nrow=d,
                        ncol=d,
                        byrow=T)
    rownames(Sigma.mat) <- asset.names
    colnames(Sigma.mat) <- asset.names

```

### Computing mean return and variance of return for random portfolio

```{r}

    #
    # randomly assigned weights (from excel worksheet)
    #
    x.vec <- c(0.794, 1.424, -1.44, 0.892, 1.03, -0.95, 0.95, -1.70)
    names(x.vec) <- asset.names


    #
    # Portfolio mean and variance
    # Note: slightly different than in-class example due to precision
    #
    mu.p <- t(x.vec) %*% mu.vec
    sigma.p <- sqrt( t(x.vec) %*% Sigma.mat %*% x.vec )

    rbind(sum=sum(x.vec),   # weights should sum to 1
          mu=mu.p[1], 
          sigma=sigma.p[1])

```

### Generate mean-variance plot for random portfolios

```{r}

    #
    # Generate a bunch of random portfolios and plot them.
    # Each column of x.mat is a random set of portfolio weights.
    # 
    set.seed(1234)
    k <- 20
    x.mat <- sapply( 1:k, 
                     function(c) {
                         x.vec <- c( runif(d-1,-2,2), 0)
                         x.vec[d] <- 1 - sum(x.vec)
                         x.vec
                     })
    rownames(x.mat) <- asset.names

    #
    # t(x.mat) %*% mu.vec 
    #   
    # returns a vector where each entry is the mean return
    # for the portfolio with the corresponding weights in x.mat
    #
    mu.p.vec <- t(x.mat) %*% mu.vec

    #
    # Loop thru x.mat and compute the variance for each col-vector
    # of portfolio weights.
    #
    sigma.p.vec <- apply(x.mat,
                         2,
                         function(x.vec) { sqrt(t(x.vec) %*% Sigma.mat %*% x.vec) })


    rbind(x.mat, 
          sum=colSums(x.mat),   # Columns (portfolio weights) should sum to 1.
          mu=t(mu.p.vec)[1,],
          sigma=sigma.p.vec)


    #
    # Plot the random portfolios
    # 
    plot(x=sigma.p.vec,
         y=mu.p.vec,
         pch=16, 
         main="Mean-Variance Plot",
         col="red",
         ylim=c(min(mu.p.vec)*2, max(mu.p.vec)*2),
         xlim=c(0, max(sigma.p.vec)*1.5), 
         xlab=expression(sigma[p]), 
         ylab=expression(mu[p]))
    abline(h=0)

```

### Use optimization (solver) to compute efficient frontier


```{r optim,cache=T,warning=F}

    library(Rsolnp)

    #
    # ----- solnp: Parameters
    #
    #     pars: The starting parameter vector.
    # 
    #      fun: The main function which takes as first argument the parameter
    #           vector and returns a single value.
    # 
    #    eqfun: (Optional) The equality constraint function returning the
    #           vector of evaluated equality constraints.
    # 
    #      eqB: (Optional) The equality constraints.
    # 
    #  ineqfun: (Optional) The inequality constraint function returning the
    #           vector of evaluated inequality constraints.
    # 
    #   ineqLB: (Optional) The lower bound of the inequality constraints.
    # 
    #   ineqUB: (Optional) The upper bound of the inequality constraints.
    # 
    #       LB: (Optional) The lower bound on the parameters.
    # 
    #       UB: (Optional) The upper bound on the parameters.
    # 
    #
    # -----  Return Value: List containing the following values:
    #
    #     pars: Optimal Parameters.
    # 
    # convergence : Indicates whether the solver has converged (0) or not (1
    #           or 2).
    # 
    #   values: Vector of function values during optimization with last one
    #           the value at the optimal.
    # 
    # lagrange: The vector of Lagrange multipliers.
    # 
    #  hessian: The Hessian of the augmented problem at the optimal solution.
    # 
    #   ineqx0: The estimated optimal inequality vector of slack variables
    #           used for transforming the inequality into an equality
    #           constraint.
    # 
    # nfuneval: The number of function evaluations.
    # 
    #  elapsed: Time taken to compute solution.
    #

    #
    # Use solnp to compute the efficient portfolio for every sigma.p
    # for the random portfolios we generated above
    #
    optimizationFn <- function(i) {
        # x.vec <- x.mat[,i]
        x.vec <- rep(1/d,d)
        sigma.p <- sigma.p.vec[i]

        sol <- solnp( pars=x.vec,
                      fun=function(x.vec) { -1 * t(x.vec) %*% mu.vec },     # (-1) because solnp minimizes this function
                      eqfun=function(x.vec) { sum(x.vec) },
                      eqB=c(1),
                      ineqfun=function(x.vec) { sqrt( t(x.vec) %*% Sigma.mat %*% x.vec ) },
                      ineqUB=sigma.p,
                      ineqLB=sigma.p-0.001,
                      control=list(trace=0))
        # warnings()

        x.eff.vec <- sol$pars
        mu.eff <- t(x.eff.vec) %*% mu.vec
        sigma.eff <- sqrt( t(x.eff.vec) %*% Sigma.mat %*% x.eff.vec )
        list(mu=mu.eff, sigma=sigma.eff, sol=sol)
    }

    mu.eff.p.vec <- rep(0,k)
    sigma.eff.p.vec <- rep(0,k)

    # for some reason sapply hangs
    # sapply( 1:k, optimizationFn)      
    for (i in 1:k) {

        l <- optimizationFn(i)
        mu.eff.p.vec[i] <- l$mu
        sigma.eff.p.vec[i] <- l$sigma
    }

```

And now plot the efficient frontier:

```{r}

    #
    # why the fltr?  cuz sometimes solnp doesn't converge, which
    # ends up producing wild results.  Just ignore those.
    #
    fltr <- abs( sigma.eff.p.vec - sigma.p.vec ) < 0.01
    mu.eff.p.vec <- mu.eff.p.vec[fltr]
    sigma.eff.p.vec <- sigma.eff.p.vec[fltr]

    #
    # Plot the random portfolios
    # 
    plot(x=sigma.p.vec,
         y=mu.p.vec,
         pch=16, 
         main="Efficient Frontier",
         col="red",
         ylim=c(min(mu.p.vec)*2, max(mu.eff.p.vec)*1.25),
         xlim=c(0, max(sigma.p.vec)*1.5), 
         xlab=expression(sigma[p]), 
         ylab=expression(mu[p]))
    abline(h=0)

    #
    # Now plot the efficient frontier
    #
    points(x=sigma.eff.p.vec,
           y=mu.eff.p.vec,
           pch=16,
           col="blue")

        
```


### Solve for efficient frontier analytically 

```{r}

    source("fin-eng-2.R")

    #
    # Global minimum variance portfolio: portfolio that has
    # the minimum variance (risk).
    #
    gmv.weights.vec <- meanVarGetMinVarPortfolio( mu.vec, Sigma.mat )

    #
    # Need a 2nd portfolio along the efficient frontier.
    # The two-fund theorem tells us that every efficient portfolio is a linear
    # combination of any two other efficient portfolios.
    #
    eff.weights.vec <- meanVarGetEfficientPortfolio( mu.vec, Sigma.mat, target.return=max(mu.eff.p.vec) )

    #
    # Use the two efficient portfolios to build the efficient frontier
    #
    eff.weights.mat <- meanVarBuildEfficientFrontier(mu.vec, 
                                                     Sigma.mat, 
                                                     alpha.vec=seq(from=-1,to=2,by=0.1), 
                                                     gmv.portfolio.weights=gmv.weights.vec, 
                                                     eff.portfolio.weights=eff.weights.vec) 

    eff.frontier.means <- apply(eff.weights.mat,
                                2,
                                function(x.vec) { t(x.vec) %*% mu.vec })

    eff.frontier.sigmas <- apply(eff.weights.mat,
                                 2,
                                 function(x.vec) { sqrt(t(x.vec) %*% Sigma.mat %*% x.vec) })

    #
    # Plot the random portfolios
    # 
    plot(x=sigma.p.vec,
         y=mu.p.vec,
         pch=16, 
         main="Efficient Frontier",
         col="red",
         ylim=c(min(mu.p.vec)*2, max(mu.eff.p.vec)*1.25),
         xlim=c(0, max(sigma.p.vec)*1.5), 
         xlab=expression(sigma[p]), 
         ylab=expression(mu[p]))
    abline(h=0)

    #
    # Plot the efficient frontier as computed by solver.
    #
    points(x=sigma.eff.p.vec,
           y=mu.eff.p.vec,
           pch=16,
           col="blue")

    #
    # Plot the efficient frontier as computed analytically.
    #
    points(x=eff.frontier.sigmas,
           y=eff.frontier.means,
           type="b",
           pch=16,
           col="green")


```

<br /><br />
## <a name="Lesson-1.3-Sharpe-Optimal-Portfolio-and-the-Risk-Free-Frontier"></a>Lesson 1.3 Sharpe Optimal Portfolio and the Risk-Free Frontier

### Sharpe Optimal Portfolio (aka Tangency Portfolio)

```{r}

    rf <- 1.5 / 100

    sharpe.optimal.weights <- meanVarGetTangencyPortfolio( mu.vec = mu.vec,
                                                           Sigma.mat = Sigma.mat,
                                                           risk.free.rate = rf )

    sharpe.optimal.port <- calcPortfolioMeanVarReturn( weights.vec = sharpe.optimal.weights,
                                                       mu.vec = mu.vec,
                                                       Sigma.mat = Sigma.mat )

    sharpe.optimal.port$sharpe.ratio <- (sharpe.optimal.port$expected.return - rf) / sharpe.optimal.port$stdev.return

    sharpe.optimal.port$weights
    cbind(sharpe.optimal.port)

```

### Security Market Line

```{r}

    #
    # Multiply the portfolio weights by the covariances of each asset with all
    # other assets (the rows of Sigma.mat).  Then divide by the variance of the 
    # market (where the market portfolio is the sharpe optimal portfolio).
    #
    beta.vec <- Sigma.mat %*% sharpe.optimal.port$weights / sharpe.optimal.port$stdev.return^2

    #
    # Plot the security market line
    # The security market line is the relationship between beta and market excess returns
    # 
    plot(x=beta.vec,
         y=beta.vec * (sharpe.optimal.port$expected.return - rf),
         pch=16, 
         type="b",
         main="Security Market Line",
         col="blue",
         # ylim=c(min(mu.p.vec)*2, max(mu.eff.p.vec)*1.25),
         # xlim=c(0, max(sigma.p.vec)*1.5), 
         xlab="beta", 
         ylab="beta * excess market return")

    #
    # Now plot the actual excess returns for each asset
    #
    points(x=beta.vec,
           y=mu.vec - rf,
           pch=16,
           col="red")

```




